


```{r}
install.packages("nnet")
install.packages("tidyverse")
install.packages("caret")
install.packages("ordinal")
```

```{r}
library(tidyverse)
library(caret)
library(nnet)
library(ordinal)
library(ggplot2)
```
```{r}
data <- read.csv("Sleep_Efficiency.csv")

```

```{r}
head(data)
```
```{r}

age<- data[,2]
gender<- data[,3]
Bedtime<- data[,4]
Wakeup_time<- data[,5]
sleep_duration<- data[,6]
sleep_efficiency<- data[,7]
REM_sleep<- data[,8]
deep_sleep<- data[,9]
light_sleep<- data[,10]
Awakenings<- data[,11]
Caffeine_consumption<- data[,12]
Alcohol_consumption<- data[,13]
Smoking_status<- data[,14]
Exercise_frequency<- data[,15]
Gender_coded<- data[,16]
Smoke_coded<- data[,17]
```


```{r}
summary(age)
summary(gender)
summary(Bedtime)
summary(Wakeup_time)
summary(sleep_duration)
summary(sleep_efficiency)
summary(REM_sleep)
summary(deep_sleep)
summary(light_sleep)
summary(Awakenings)
summary(Caffeine_consumption)
summary(Alcohol_consumption)
summary(Smoking_status)
summary(Exercise_frequency)
```

#These are the summary statistics for each variable in the dataset:

age: The minimum age is 9, the maximum age is 69, and the mean age is 40.29.
gender: There are 452 values in this variable, all of which are character strings.

Bedtime: The minimum bedtime is 44199, which corresponds to approximately 12:16 AM, and the maximum bedtime is 44562, which corresponds to approximately 12:26 AM. The median bedtime is 44398, which corresponds to approximately 11:59 PM.

Wakeup_time: The minimum wakeup time is 44199, which corresponds to approximately 12:16 AM, and the maximum wakeup time is 44561, which corresponds to approximately 12:26 AM. The median wakeup time is 44398, which corresponds to approximately 11:59 PM.

sleep_duration: The minimum sleep duration is 5 hours, the maximum sleep duration is 10 hours, and the mean sleep duration is 7.466 hours.

sleep_efficiency: The minimum sleep efficiency is 0.5, the maximum sleep efficiency is 0.99, and the mean sleep efficiency is 0.7889.

REM_sleep: The minimum REM sleep is 15%, the maximum REM sleep is 30%, and the mean REM sleep is 22.62%.

deep_sleep: The minimum deep sleep is 18%, the maximum deep sleep is 75%, and the mean deep sleep is 24.56%.

light_sleep: The minimum light sleep is 7%, the maximum light sleep is 63%, and the mean light sleep is 52.82%.

Awakenings: The minimum number of awakenings is 0, the maximum number of awakenings is 4, and the mean number of awakenings is 1.641.

Caffeine_consumption: The minimum caffeine consumption is 0 mg, the maximum caffeine consumption is 200 mg, and the mean caffeine consumption is 23.65 mg. There are 25 missing values in this variable.

Alcohol_consumption: The minimum alcohol consumption is 0 units, the maximum alcohol consumption is 5 units, and the mean alcohol consumption is 1.791 units. There are 14 missing values in this variable.

Smoking_status: There are 452 values in this variable, all of which are character strings.

Exercise_frequency: The minimum exercise frequency is 0 days per week, the maximum exercise frequency is 5 days per week, and the mean exercise frequency is 1.174 days per week. There are 6 missing values in this variable.




Data visualization 












```{r}
#age
hist(age, col = "blue")
plot(age, col = "blue")
```

The histogram and plot show the distribution of ages in the dataset. The histogram shows the frequency of each age group on the y-axis and the age values on the x-axis. The plot shows the same information in a different way, with each dot representing an individual's age.

From the plot and histogram, we can see that the distribution of ages is roughly bell-shaped, with a mean age of around 40 years old. The majority of the individuals in the dataset are between 25 and 60 years old, with a few individuals in their 70s. There are no outliers in the dataset, as all the ages fall within a reasonable range.

```{r}
ggplot(data = data, aes(x = Gender, y = Sleep.efficiency, fill = Gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#FFCC99", "#66B2FF")) +
  labs(title = "Sleep Efficiency by Gender") +
  xlab("Gender") +
  ylab("Sleep Efficiency")



```

The boxplot shows the distribution of sleep efficiency for each gender category, where the orange box represents females and the blue box represents males. 
From the plot, it appears that on average, females have a slightly higher sleep efficiency than males. 
The median sleep efficiency for females is around 0.8, while for males it is around 0.75. 
Additionally, the distribution of sleep efficiency appears to be more tightly clustered for females than for males, with fewer outliers.

```{r}

#bed time


ggplot(data = data, aes(x = Bedtime, fill = ..count..)) +
  geom_histogram(color = "black", binwidth = 15) +
  labs(title = "Bedtime Distribution") +
  xlab("Bedtime") +
  ylab("Count")
```
The bedtime values are represented on the x-axis, and the number of observations that fall into each bedtime bin is shown on the y-axis. The histogram reveals that most people fall asleep between 22:00 and 1:00, with the peak time being about 23:00. A modest right-skew in the distribution, which is essentially symmetrical and shows that some people stay up later than others, is present.
```{r}
#wake time
ggplot(data = data, aes(x = Wakeup_time, fill = ..count..)) +
  geom_histogram() +
  scale_fill_gradient(low = "maroon", high = "pink") +
  labs(title = "Wakeup Time Distribution", y = "Count") +
  xlab("Wakeup Time")
```

```{r}
#sleep durations
# create a density plot of sleep duration data
ggplot(data = data, aes(x = sleep_duration, y = ..density.., fill = Gender)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("cyan", "magenta")) +
  labs(title = "Sleep Duration Distribution by Gender") +
  xlab("Sleep Duration") +
  ylab("Density")




```
The plot shows the distribution of sleep duration for the participants . 
The x-axis represents the sleep duration in hours, and the y-axis represents the density of observations. 
The plot is skewed to the right, indicating that the majority of participants had a sleep duration of around 7 hours.
The density curve is smooth and bell-shaped, indicating that the data follows a normal distribution. This suggests that the sleep duration for the participants is relatively consistent and predictable.







```{r}
#sleep efficiency
ggplot(data = data, aes(x = sleep_efficiency)) +
  geom_histogram(color = "black", fill = "#CCFFCC", bins = 30) +
  labs(title = "Sleep Efficiency Distribution", x = "Sleep Efficiency", y = "Frequency")
```

```{r}
# rem sleep
# rem sleep
library(ggplot2)

ggplot(data = data, aes(x = REM_sleep, y = ..density.., fill = "REM Sleep")) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("purple")) +
  labs(title = "Distribution of REM Sleep", x = "REM Sleep (minutes)", y = "Density")
```
The graph shows the distribution of REM sleep percentages among the participants. The majority of the participants have REM sleep percentages between 15 and 25, with a peak around 20. The distribution is roughly symmetric and bell-shaped, suggesting a normal distribution.

```{r}
#deep sleep
ggplot(data, aes(x = deep_sleep)) +
  geom_histogram(binwidth = 5, fill = "#66CC00", color = "black") +
  labs(title = "Distribution of Deep Sleep", x = "Deep Sleep (%)", y = "Count")


```
The histogram displays the distribution of deep sleep percentages among the participants .

The x-axis denotes the percentage of time spent in deep sleep, and the y-axis shows the count of participants 

The plot highlights that the highest number of participants spent around 10-20% of their total sleep time in deep sleep.
A small group of participants, however, experienced more than 30% of their sleep time in deep sleep. The graph has a slightly right-skewed distribution, with a longer tail towards the right-hand side, signifying that a few participants had a higher percentage of deep sleep than the majority.
```{r}
#light sleep
ggplot(data, aes(x = light_sleep)) +
  geom_histogram(binwidth = 5, fill = "#E7E9C1", color = "black") +
  labs(title = "Distribution of Deep Sleep", x = "Deep Sleep (%)", y = "Count")
```
This histogram shows the distribution of light sleep percentages among the participants . The x-axis represents the percentage of time spent in light sleep, while the y-axis represents the count or number of participants who fall into each bin. The bin width is 5, so the histogram groups light sleep percentages into intervals of 5.

From the graph, we can see that the majority of participants spent between 50-65% of their total sleep time in light sleep.
There is a small group of participants who spent less than 40% of their total sleep time in light sleep. The distribution is slightly left-skewed, with a longer tail on the left-hand side of the graph, indicating that a few participants had a lower percentage of light sleep than the majority.
```{r}
#awake
library(ggplot2)

ggplot(data, aes(x = Awakenings)) +
  geom_histogram(aes(y = ..density..), 
                 binwidth = 0.5, 
                 fill = "lightblue", 
                 color = "black", 
                 alpha = 0.6) +
  geom_density(fill = "pink", alpha = 0.6) +
  labs(title = "Awakenings Distribution", y = "Density") +
  xlab("Number of Awakenings per Night")
```
The histogram shows the distribution of the number of awakenings during the night's sleep among the sample population. 
The x-axis represents the number of awakenings and the y-axis represents the frequency or count of participants who had that many awakenings. The majority of participants had 0-5 awakenings, with a peak at around 1 awakening, while very few participants had more than 10 awakenings. 
This suggests that most people in the sample had a relatively undisturbed sleep with minimal interruptions, while a small percentage of people had a more disrupted sleep pattern with several awakenings throughout the night.
```{r}
#caffine 
barplot(table(Caffeine_consumption), col = "#69B3A2", xlab = "Caffeine Consumption", ylab = "Count", main = "Distribution of Caffeine Consumption")

```
The bar plot shows the distribution of caffeine consumption among the participants in the dataset. The x-axis represents the different levels of caffeine consumption, while the y-axis represents the count or number of participants falling into each category. The categories range from 0-1 cups per day to 7 or more cups per day.

From the graph, we can see that the majority of participants consumed between 1-3 cups of caffeine per day. There is a gradual decrease in the number of participants who consumed higher levels of caffeine, with very few participants consuming more than 5 cups per day. The highest number of participants falls into the 1-2 cups per day category. This bar plot gives a clear and concise visual representation of the caffeine consumption distribution in the dataset.

```{r}
#alchol

ggplot(data, aes(x = Alcohol_consumption)) + 
  geom_histogram(binwidth = 1, fill = "cyan", color = "black") +
  labs(title = "Frequency of Alcohol Percentage", x = "Alcohol Percentage", y = "Count")


```
The histogram represents the frequency distribution of alcohol consumption percentages among participants in the dataset. The x-axis shows the range of alcohol percentages, with a bin width of 1. The y-axis displays the count or number of participants that fall into each bin.

From the graph, it is clear that the majority of participants consumed alcohol in the range of 0-5%, with a sharp decrease in the frequency of consumption as the alcohol percentage increases. Only a few participants consumed alcohol at higher percentages, with the highest percentage being around 17%. The distribution is slightly skewed to the right, indicating that a few participants consumed higher percentages of alcohol than the majority.
```{r}
#smoke cod
barplot(table(ifelse(Smoke_coded == 1, "Yes", "No")), col = "#69B3A2", xlab = "Smoking Status", ylab = "Count", main = "Distribution of Smoking Status")
```
From the graph, we can see that the majority of participants did not smoke, as indicated by the taller "No" bar on the left. There is a smaller group of participants who did smoke, as indicated by the shorter "Yes" bar on the right.
```{r}
#exercise
ggplot(data, aes(x=Exercise_frequency)) +
  geom_density(fill="lightblue", alpha=0.6) +
  geom_rug() +
  labs(title = "Exercise Frequency Distribution", y = "Density") +
  xlab("Exercise Frequency")




```
The densityplot shows the distribution of exercise frequency among the sample population. 
The x-axis represents the number of times participants engage in exercise per week, while the y-axis shows the density of observations at each value of x. 
The density plot indicates that the most common frequency of exercise is between 1-3 times per week, and the distribution is skewed to the right, suggesting that fewer participants exercise frequently (i.e., more than 3 times per week).












#modelsss



```{r}
mod1=lm(age~sleep_efficiency)#regression between age and sleep efficiency
summary(mod1)
```
This output shows , where the dependent variable is age and the independent variable is sleep_efficiency. The model estimates that the intercept value is 32.728 and the coefficient for sleep_efficiency is 9.580.

The standard error of the coefficient estimates and the t-value for the coefficient are also provided. The p-value for the coefficient is 0.0366, which is less than 0.05, indicating that there is a significant relationship between age and sleep efficiency.

The R-squared value is 0.0097, which means that only 0.97% of the variance in age is explained by sleep efficiency. The adjusted R-squared value is slightly lower at 0.0075, which takes into account the number of independent variables in the model.

The residual standard error is 13.12, which is the estimate of the standard deviation of the errors in the model. It suggests that the model has a moderate amount of variability that is not explained by sleep efficiency.

Overall, the model suggests that sleep efficiency has a significant, but weak positive relationship with age.


```{r}
# Subset the data by gender
male_data <- subset(data, Gender == "Male")
```


```{r}
head((male_data))
```




```{r}
female_data <- subset(data, Gender == "Female")
```

```{r}


head(female_data)
```


```{r}
#male
male_mean <- mean(male_data$Deep.sleep.percentage)
male_sd <- sd(male_data$Deep.sleep.percentage)

male_mean

male_sd
```



```{r}
#female
female_mean <- mean(female_data$Deep.sleep.percentage)
female_sd <- sd(female_data$Deep.sleep.percentage)


female_mean

female_sd
```


```{r}
#t test
t_test_result <- t.test(male_data$Deep.sleep.percentage, female_data$Deep.sleep.percentage, var.equal = TRUE)

# Print the t-test results
print(t_test_result)
```

This two-sample-test is  comparing the mean values of Deep.sleep.percentage between two groups: male_data and female_data. The null hypothesis is that the true difference in means between the two groups is equal to 0. The alternative hypothesis is that the true difference in means is not equal to 0.

The t-value is 1.6156 with 450 degrees of freedom, and the p-value is 0.1069. Since the p-value is greater than 0.05, we fail to reject the null hypothesis at the 5% significance level, and we conclude that there is not enough evidence to suggest that the mean values of Deep.sleep.percentage between males and females are different.

The 95% confidence interval for the difference in means is -0.5140398 to 5.2640398. This interval includes 0, which supports our conclusion that there is no statistically significant difference between the two groups.

The mean value of Deep.sleep.percentage for males is 54.000 and for females is 51.625, based on the sample estimates.


```{r}
mod2=lm(Caffeine_consumption~Awakenings)#regression between the number of awakenings and caffeine consumption
summary(mod2)
mod3=lm(sleep_duration~Exercise_frequency)#regression between the exercise frequency and sleep duration
summary(mod3)
mod4=lm(sleep_duration~Caffeine_consumption+Alcohol_consumption)#regression between caffeine cons, alcohol cons, and sleep duration
summary(mod4)
```

These results show about linear regression models, each predicting a dependent variable (Caffeine_consumption, sleep_duration) based on one or two independent variables (Awakenings, Exercise_frequency, Caffeine_consumption, Alcohol_consumption).

For the first model, the equation is: Caffeine_consumption = 27.288 - 2.415 * Awakenings. The coefficient for Awakenings is significant (p-value = 0.0283), indicating that there is a negative linear relationship between caffeine consumption and the number of awakenings during the night.

For the second model, the equation is: sleep_duration = 7.53508 - 0.04148 * Exercise_frequency. The coefficient for Exercise_frequency is not significant (p-value = 0.15), indicating that there is not a significant linear relationship between sleep duration and exercise frequency.

For the third model, the equation is: sleep_duration = 7.5115482 - 0.0006453 * Caffeine_consumption - 0.0207350 * Alcohol_consumption. Neither of the independent variables (Caffeine_consumption and Alcohol_consumption) are significant (p-value = 0.656 and p-value = 0.445, respectively), indicating that there is not a significant linear relationship between sleep duration and either caffeine or alcohol consumption.

The "Multiple R-squared" values for all three models are quite low, indicating that the models explain only a small proportion of the variance in the dependent variable. The "Adjusted R-squared" values are slightly lower than the "Multiple R-squared" values, indicating that adding additional independent variables to the models did not significantly improve the fit.








```{r}
# Subset the data by smoking status
REM_smoke <- subset(data, Smoke == "1")


print(REM_smoke)
```


```{r}
#no smoke 
#print(REM_smoke)
REM_No_smoke <- subset(data, Smoke == "0")

print(REM_No_smoke)
```


```{r}
# Calculate the mean and standard deviation of REM sleep percentage for each group
REM_smoke_mean <- mean(REM_smoke$REM.sleep.percentage)
REM_smoke_sd <- sd(REM_smoke$REM.sleep.percentage)
REM_No_smoke_mean <- mean(REM_No_smoke$REM.sleep.percentage)
REM_No_smoke_sd <- sd(REM_No_smoke$REM.sleep.percentage)







REM_smoke_mean 
REM_smoke_sd 
REM_No_smoke_mean 
REM_No_smoke_sd


```





```{r}
#t test
t_test_result1 <- t.test(REM_smoke$REM.sleep.percentage, REM_No_smoke$REM.sleep.percentage, var.equal = TRUE)
```



```{r}
# Print the t-test results
print(t_test_result1)
mean_bedtime=mean(Bedtime)
mean_wakeup_time=mean(Wakeup_time)
print(mean_bedtime)
print(mean_wakeup_time)

mod5=lm(age~Awakenings)#regression between age and the number of awakenings
summary(mod5)
mod6=lm(sleep_efficiency~Exercise_frequency)#regression between sleep efficiency and exercise frequency
summary(mod6)
mod7=lm(REM_sleep~deep_sleep+light_sleep)#regression between REM sleep percentage, deep sleep percentage, and light sleep percentage
summary(mod7)

```
The t-test result for the comparison of REM sleep percentage between smokers and non-smokers showed a p-value of 0.4949, indicating that there is no significant difference in the mean REM sleep percentage between the two groups. The mean REM sleep percentage for smokers was 22.77, and for non-smokers, it was 22.53.

The regression analysis between age and the number of awakenings showed that there is no significant relationship between these two variables (p-value of 0.7124), and the model explains only 0.03% of the variation in the age variable.

The regression analysis between sleep efficiency and exercise frequency showed a significant positive relationship (p-value of 2.669e-08). However, the warning "essentially perfect fit" suggests that the model may not be reliable.

The regression analysis between REM sleep percentage, deep sleep percentage, and light sleep percentage showed a perfect fit, with R-squared and adjusted R-squared both equal to 1. This means that the model can explain 100% of the variation in the REM sleep variable using deep sleep and light sleep variables. However, this perfect fit is unlikely in real-world scenarios and may suggest an issue with the data or the model.


